---
title: "03-2023-pilot-ni-pre-exposure-cells"
subtitle: "Pre-exposure to nickel effects on Primary Epithelial Cells of *Botryllus schlosseri*"
output: html_document
date: "2023-12-24"
---

```{r, warning=FALSE}
library(knitr)
library(tidyverse)
library(tidyr)
library(dplyr)
library(hrbrthemes)
library(ggplot2)
library(car)
library(RColorBrewer)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE)
```

# Objective and Background

Data below was collected from a pilot study conducted in Autumn 2023 where field-collected individual *Botryllus schlosseri* were exposed to a low and high treatment of nickel chloride and then were dissected for primary buds. The objective was to evaluate if a pre-nickel exposure in *Botryllus* improved primary epithelial cell monolayer proliferation. Refer to [Pilot 1.1 Nickel Exposure](https://valeste.github.io/tough-tunicates/posts/nicl2-botryllus-pilot/) notebook post for experimental design details.

# Retrieving Data from Google Sheets

Note, if you do this, make sure your spreadsheet is publicly available to anyone with a link.

```{r, engine='bash', eval= FALSE, echo=TRUE}
cd ..
curl -L https://docs.google.com/spreadsheets/d/13o7wsaFG9QAbBtIo31XtbfPtFOcf5HUEEYDksrcqaok/export?exportFormat=csv | tee data/ni_pre-exposure_cell.csv
```

```{r, engine='bash', eval= FALSE, echo=TRUE}
cd ..
curl -L https://docs.google.com/spreadsheets/d/1jscIxXDsEARLXBi_zoA6q5WON_mEWq42oqZWRZD7GHg/export?exportFormat=csv | tee data/ni_pre-exposure_cell_18hps_cellcounts.csv
```


Read in the data to your local R environment.

```{r}
cell <- read.csv(file = "../data/ni_pre-exposure_cell.csv")  
cell18 <- read.csv(file ="../data/ni_pre-exposure_cell_18hps_cellcounts.csv")
```

# Cleaning up Data

During the study, we documented blastogenic stage with the 7 level blastogenic stage scale For simplicity in our graphs and our interpretation, we will be being the stages into a 4 level blastogenic stage scale

```{r}
# Create a new column 'simple_stage' based on conditions
cell <- cell %>%
  mutate(simple_stage = case_when(
    stage_at_dissection %in% c("A1", "A2") ~ "A",
    stage_at_dissection %in% c("B1", "B2") ~ "B",
    stage_at_dissection %in% c("C1", "C2") ~ "C",
    stage_at_dissection == "TO" ~ "TO",
    TRUE ~ NA_character_  # This will handle any other cases or return NA if none match
  ))

cell18 <- cell18 %>%
  mutate(simple_stage = case_when(
    stage_at_dissection %in% c("A1", "A2") ~ "A",
    stage_at_dissection %in% c("B1", "B2") ~ "B",
    stage_at_dissection %in% c("C1", "C2") ~ "C",
    stage_at_dissection == "TO" ~ "TO",
    TRUE ~ NA_character_  # This will handle any other cases or return NA if none match
  ))

```

Cleaning up and subsetting data.

## for just the instances of monolayerss
```{r}
cell$date <- mdy(cell$date) #convert the date column from characters to true date

cell_pilot <- cell[cell$experiment == 0, ] # only want the pilot experiment, experiment 1 failed in poor dissecting technique

# Subset the 'cell_pilot' data frame to include rows with hours post seeding (hps) values of 0, 18, 38, and 70 hps.

cell_pilot <- cell_pilot[cell_pilot$hps %in% c(0, 18, 38, 70, 240, 600), ]

# Set the treatment, animal ID, and date as factors.
cell_fact <- cell_pilot %>%
  mutate(treatment_mg_per_L = as.factor(treatment_mg_per_L)) %>%
  mutate(animal_ID = as.factor(animal_ID)) %>%
  mutate(date = as.factor(date))


summary(cell_fact)
```


## for with cell counts at 18hps
```{r}
cell18$date <- mdy(cell18$date) #convert the date column from characters to true date



# Set the treatment, animal ID, and date as factors.
cell_fact18 <- cell18 %>%
  mutate(treatment_mg_per_L = as.factor(treatment_mg_per_L)) %>%
  mutate(animal_ID = as.factor(animal_ID)) %>%
  mutate(date = as.factor(date))


summary(cell_fact18)
```

# calculate percent output monolayer instances
Calculate the percent of monolayer output per seeded tissue piece based on the first 12 rows of 'attached_tissue'. Here we are calculating the amount of monolayers that occurred relative to the amount of originally seeded tissue. Overtime, seeded tissue detaches and is removed from the plate so if you were to calculate this ratio per day it would be off. If using this, make sure the data frame is in order from A1-C4 wells otherwise it's miscalculated.

```{r}
percent_output <- ((cell_fact$no_mono_out) / head(cell_fact$no_attached_tissue, 12)) * 100

# Add the newly calculated percent output values to the old data frame
cell_fact <- cbind(cell_fact,percent_output)

```

# Summary Statistics for just cell data frame. calculating monolayer output by percent instance

Number of tissue pieces attached:


## Number of monolayer output:

```{r}
# Calculate the mean 
mean_mono_out <- cell_fact %>%
  group_by(hps, treatment_mg_per_L, simple_stage) %>%
  summarise(no_mono = mean(no_mono_out, na.rm = TRUE),
            num_points = n())


# Calculate SD and filter out groups with fewer than 2 observations
SD_mono_out <- cell_fact %>%
  group_by(hps, treatment_mg_per_L, simple_stage) %>%  
  summarise(SD_mono = sd(no_mono_out, na.rm = TRUE),
            num_points = n())


n <- 4

SE_mono <- SD_mono_out$SD_mono/ sqrt(n)
# Print the calculated mean health scores
print(mean_mono_out)
print(SE_mono)
```

## Monolayer output by relative tissue seeded (percent):

```{r}
# Calculate the mean 
mean_ratio_mono_out <- cell_fact %>%
  group_by(hps, treatment_mg_per_L, simple_stage) %>%
  summarise(percent_mean_mono = mean(percent_output, na.rm = TRUE),
            num_points = n())


# Calculate SD and filter out groups with fewer than 2 observations
SD_ratio_mono_out <- cell_fact %>%
  group_by(hps, treatment_mg_per_L, simple_stage) %>%  
  summarise(SD_ratio_mono = sd(percent_output, na.rm = TRUE),
            num_points = n())


n <- 4

SE_mono_ratio <- SD_ratio_mono_out$SD_ratio_mono/ sqrt(n)
# Print the calculated mean health scores
print(mean_ratio_mono_out)
print(SE_mono_ratio)
```


# Graphs for monolayer output (first iteration for SICB 2023)

Below are graphs used in the SICB poster generated. Graph settings are set to be really large for printing and don't render well in this format so instead are figures with attached images of the produced graphs.

```{r, eval=FALSE}
setwd('..')
png(filename = "output/line_percnet_output.png", width = 1100, height = 700) 
##good for making a 7x5 in image for printing ~150 dpi

ggplot(data = mean_ratio_mono_out, aes(x = hps, y = percent_mean_mono, 
                                       color = simple_stage, linetype = treatment_mg_per_L)) +
  theme_minimal() +
  labs(x = "Hours Post-Seeding", y = "Percent Monolayer Outputs") +
  geom_point() +
  geom_line(size = 3) +  # Adjusted line thickness to 2
  scale_color_manual(name = "Stage",
                     values = c("#FFD92F", "#E78AC3",  "#8DA0CB", "#A6D854"),  # Specify the colors
                     labels = c("A", "B", "C", "TO")) +  # Use fill instead of color in the legend
  scale_linetype_manual(name = "Treatment",
                        values = c("dotted", "dotdash", "solid"),  # Example linetypes
                        labels = c("Control", "5 mg/L", "45 mg/L")) +
  scale_x_continuous(breaks = seq(0, max(mean_ratio_mono_out$hps), by = 10), expand = c(0, 0)) +  # Adjust x-axis to start at 0
  scale_y_continuous(expand = c(0, 0), breaks = c(0, 5, 10, 15, 20, 25), limits = c(0, 25)) +  # Adjust y-axis to start at 0
  theme(legend.position = "right",
        legend.margin = margin(1, 1, 1, 50),  # Adjusting the margin around the legend
        axis.text = element_text(size = 30),
        axis.title = element_text(size = 40),
        legend.text = element_text(size = 30),
        legend.title = element_text(size = 40),
        plot.margin = margin(t = 2, r = 1, b = 1, l = 1, unit = "cm"),  # Adjust the top margin
        legend.key.size = unit(4, "lines"),
        panel.grid.major.x = element_line(color = "darkgrey", size = 0.5),  # Remove major vertical gridlines
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "darkgrey", size = 0.5),
        panel.grid.minor.y = element_blank())  # Adjust the size of legend keys for linetypes

dev.off()
```




```{r}
setwd('..')
png(filename = "output/cell_attached_stacked_all_simple_stage.png", width = 1000, height = 900)

ggplot(mean_tissue_attached_18hps, aes(x = treatment_mg_per_L, y = mean_attached, fill = simple_stage)) +
   geom_bar(stat = "identity", width = 0.9) +
   scale_fill_manual(name = "Stage",
                     values = c("#FFD92F", "#E78AC3",  "#8DA0CB", "#A6D854"),  # Specify the colors
                     labels = c("A", "B", "C", "TO")) +
   labs(x = "Treatment (Nickel Chloride mg/L)",
        y = "Number of Attached Tissue") +
   theme_minimal() +
   scale_y_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30), limits = c(0, 30)) +
   theme(panel.grid.major.x = element_blank(),
         panel.grid.minor.x = element_blank(),
         panel.grid.major.y = element_line(color = "darkgrey", size = 1),
         panel.grid.minor.y = element_line(color = "darkgrey", size = 0.5), 
         axis.text = element_text(size = 30),
         axis.title = element_text(size = 40),
         legend.text = element_text(size = 40),
         legend.title = element_text(size = 40))
dev.off()
```


# Refined singular figure for monolayer instnaces thesis paper

first write out a csv of the newly formatted data frame that includes simple stage
```{r}
cell_fact
write.csv(cell_fact, "../output/cell_fact.csv", row.names = FALSE)
```


here we also look at 18 hps because that is the time point at which we see cells
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# --- Choose the timepoints you want to plot --- 
# I chose 18 hps because that is when you first start seeing monolayers
attach_hps <- 18

# choose the timepoint you want to summarize at
hps_use <- 18  

# define the number of genets represented in each stage x treatment combo
genet_n <- cell_fact %>%
  filter(hps == hps_use) %>%
  group_by(treatment_mg_per_L, simple_stage) %>%
  summarise(
    n_genets = n_distinct(animal_ID),
    .groups = "drop"
  )


```

```{r}
# This will clean up our data cell_fact clearly to use it for plotting stacked bar graphs of the outcomes for seeded tissue pieces.
option1_wide <- cell_fact %>%
  filter(hps == hps_use) %>%
  mutate(
    seeded_n  = tissue_total,
    attached_n = no_attached_tissue,
    mono_n    = no_mono_out,

    not_attached_n = pmax(seeded_n - attached_n, 0),
    attached_no_mono_n = pmax(attached_n - mono_n, 0),
    monolayer_n = pmax(mono_n, 0)
  ) %>%
  group_by(treatment_mg_per_L, simple_stage) %>%
  summarise(
    not_attached = mean(not_attached_n, na.rm = TRUE),
    attached_no_mono = mean(attached_no_mono_n, na.rm = TRUE),
    monolayer = mean(monolayer_n, na.rm = TRUE),
    n_wells = n(),
    .groups = "drop"
  ) %>%
  mutate(
    not_attached = ceiling(not_attached),
    attached_no_mono = ceiling(attached_no_mono),
    monolayer = ceiling(monolayer)
  )

```

```{r}
option1_long <- option1_wide %>%
  tidyr::pivot_longer(
    cols = c(not_attached, attached_no_mono, monolayer),
    names_to = "outcome",
    values_to = "count"
  ) %>%
  mutate(
    outcome = factor(
      outcome,
      levels = c("not_attached", "attached_no_mono", "monolayer"),
      labels = c("Not attached", "Attached, no monolayer", "Monolayer formed")
    )
  )
# create object for defining genet number in stacked bar plot
n_labels <- option1_long %>%
  left_join(genet_n,
            by = c("treatment_mg_per_L", "simple_stage")) %>%
  filter(outcome == "Monolayer formed") %>% # this is just to make sure we have only one label per stage x treatment superimposed on graph
  mutate(label = paste0("n = ", n_genets))


```

```{r}
png(filename = "../output/option1_counts_stage_stacked_facet_treatment.png", width = 1300, height = 850)

ggplot(option1_long, aes(x = simple_stage, y = count, fill = outcome)) +
  geom_col(width = 0.85) +
  facet_wrap(~ treatment_mg_per_L,nrow = 1,
  labeller = labeller(
    treatment_mg_per_L = function(x) paste0(x)
  )
) +
  scale_fill_manual(
    values = c(
      "Not attached" = "snow2",
      "Attached, no monolayer" = "snow3",
      "Monolayer formed" = "snow4"),
    name = "Total Tissue Outcomes and Cell Monolayer Output",
    labels = c(
      "Not attached" = "Tissue not attached", 
      "Attached, no monolayer" = "Tissue attached, no monolayer",
      "Monolayer formed" = "Tissue attached, monolayer formed")
  ) +
  labs(
    x = "Stage",
    y = "Number of zooid and bud tissue pieces"
  ) +
  geom_text(
    data = n_labels,
    aes(x = simple_stage, y = 0, label = label),
    vjust = 1.5,
    size = 7,
    inherit.aes = TRUE
  ) +
   scale_y_continuous(
    breaks = 0:10,
    limits = c(0, 10),
    expand = c(0, 0.7)
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect( color = "black"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey",linetype = "dashed"),
    panel.grid.minor.y = element_blank(),
    axis.text = element_text(size = 24),
    axis.title = element_text(size = 32),
    strip.text = element_text(size = 28, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    legend.title = element_text(size = 24, face = "bold"),
    legend.text = element_text(size = 24),
    legend.position = "bottom",
    legend.title.position = "top"
  )

dev.off()
```

```{r}
option2_wide <- cell_fact %>%
  filter(hps == hps_use) %>%
  mutate(
    float_zooid  = float_zooid,
    float_bud = float_prim,
    
    attached_zooid = pmax(zooid_no - float_zooid, 0),
    attached_bud  = pmax(prim_bud_no - float_prim, 0)
  ) %>%
  group_by(treatment_mg_per_L, simple_stage) %>%
  summarise(
    not_attached_z = mean(float_zooid, na.rm = TRUE),
    attached_z = mean(attached_zooid, na.rm = TRUE),
    not_attached_p = mean(float_bud, na.rm = TRUE),
    attached_p = mean(attached_bud, na.rm = TRUE),
    n_wells = n(),
    .groups = "drop"
  ) %>%
  mutate(
    not_attached_z = ceiling(not_attached_z),
    attached_z = ceiling(attached_z),
    not_attached_p = ceiling(not_attached_p),
    attached_p = ceiling(attached_p)
  )

```

```{r}
option2_long <- option2_wide %>%
  tidyr::pivot_longer(
    cols = c(not_attached_z, attached_z, not_attached_p, attached_p),
    names_to = "outcome",
    values_to = "count"
  ) %>%
  mutate(
    outcome = factor(
      outcome,
      levels = c("not_attached_z", "not_attached_p", 
                 "attached_z", "attached_p"),
      labels = c("Unattached Zooids", "Unattached Primary Buds",
                 "Attached Zooids", "Attached Primary Buds")
    )
  )
```

```{r}
png(filename = "../output/option2_counts_stage_stacked_facet_treatment.png", width = 1300, height = 850)

ggplot(option2_long, aes(x = simple_stage, y = count, fill = outcome)) +
  geom_col(width = 0.85) +
  facet_wrap(~ treatment_mg_per_L,nrow = 1
  #labeller = labeller(
    #treatment_mg_per_L = function(x) paste0(x, " mg/L NiCl2") this would give you labeling with # pasted mg/L after
  )
 +
  scale_fill_manual(
    values = c(
      "Unattached Zooids" = "lightsteelblue1",
      "Unattached Primary Buds" = "lightsteelblue2",
      "Attached Zooids" = "lightsteelblue3",
      "Attached Primary Buds"= "lightsteelblue4"),
    name = "Tissue Type Attachment Outcomes"
  ) +
  labs(
    x = "Stage",
    y = "Number of zooid or bud tissue pieces"
  ) +
  geom_text(
    data = n_labels,
    aes(x = simple_stage, y = 0, label = label),
    vjust = 1.5,
    size = 7,
    inherit.aes = TRUE
  ) +
   scale_y_continuous(
    breaks = 0:10,
    limits = c(0, 10),
    expand = c(0, 0.7)
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect( color = "black"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey",linetype = "dashed"),
    panel.grid.minor.y = element_blank(),
    axis.text = element_text(size = 24),
    axis.title = element_text(size = 32),
    strip.text = element_text(size = 28, face = "bold"),
    strip.background = element_rect(fill = "grey"),
    legend.title = element_text(size = 24, face = "bold"),
    legend.text = element_text(size = 24),
    legend.position = "bottom",
    legend.title.position = "top"
  )

dev.off()
```





# Refined singular figure(s) for cell count 

```{r}
mono_by_tissue <- cell_fact18 %>%
  mutate(
    treatment_mg_per_L = as.factor(treatment_mg_per_L),
    simple_stage = factor(simple_stage, levels = c("A","B","C","TO")),
    under_tissue = ifelse(is.na(under_tissue.), "NA", under_tissue.)
  ) %>%
    filter(no_mono_out == 1) %>%   # only wells where monolayer formed

  group_by(tissueWithCells_ID, well_Placements, animal_ID,
           treatment_mg_per_L, simple_stage) %>%
  summarise(
    cell_count_monolayer = sum(cell_count, na.rm = TRUE),  # ADD counts across images
    n_images = sum(!is.na(cell_count)),
    
    under_tissue = dplyr::coalesce(first(under_tissue[!is.na(under_tissue)]), "NA")
,
    
    .groups = "drop"
  )

```


```{r}

# 1) Collapse to "one row = one monolayer (one well)"
mono_well <- mono_by_tissue %>%
  group_by(well_Placements, animal_ID, treatment_mg_per_L, simple_stage) %>%
  summarise(
    cell_count_per_monolayer = sum(cell_count_monolayer),
    under_tissue = first(under_tissue),
    .groups = "drop"
  )


# 2) n labels per facet x stage (number of wells/monolayers)
n_lab_mono <- mono_well %>%
  count(treatment_mg_per_L, simple_stage, name = "n_wells")


```



```{r}
png(filename = "../output/cell_count_18hps_preni_mono_well.png", width = 1000, height = 800)
ggplot(mono_well,
       aes(x = simple_stage,
           y = cell_count_per_monolayer,
           fill = under_tissue)) +

  geom_col(width = 0.85, color = "white") + 

  facet_wrap(~ treatment_mg_per_L, nrow = 1) +

  scale_fill_manual(
    values = c(
      "yes" = "lightsteelblue3",
      "no"  = "lightsteelblue4",
      "NA"  = "grey50"
    ),
    labels = c(
      "yes" = "Emerging from\nunderneath tissue explant",
      "no"  = "Whole monolayer",
      "NA"  = "No monolayer"
    ),
    name = "Monolayer Cell Count Area"
  ) +

  labs(
    x = "Blastogenic Stage",
    y = "Cell count per monolayer instance observed"
  ) +

  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +

  geom_text(
    data = n_lab_mono,
    aes(x = simple_stage, y = -6, label = paste0("n = ", n_wells)),
    inherit.aes = FALSE,
    vjust = 2,
    size = 6
  ) +

  coord_cartesian(clip = "off") +

  theme_classic(base_size = 18) +
  theme(
    strip.text = element_text(size = 22, face = "bold"),
    axis.title = element_text(size = 22),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14),
    legend.position = "right",
    panel.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA)
  )

dev.off()
```




```{r}
well_counts <- cell_fact18 %>%
  mutate(
    treatment_mg_per_L = as.factor(treatment_mg_per_L),
    simple_stage = factor(simple_stage, levels = c("A","B","C","TO")),
    under_tissue = ifelse(is.na(under_tissue.), "NA", under_tissue.)
  ) %>%
  group_by(well_Placements, animal_ID, treatment_mg_per_L, simple_stage) %>%
  summarise(
    cell_count_per_well = sum(cell_count, na.rm = TRUE),
    under_tissue = dplyr::coalesce(first(under_tissue[!is.na(under_tissue)]), "NA"),
    n_images = sum(!is.na(cell_count)),
    .groups = "drop"
  )

n_lab_genets <- well_counts %>%
  count(treatment_mg_per_L, simple_stage, name = "n_genets")
n_lab_well <- mono_well %>%
  count(treatment_mg_per_L, simple_stage, animal_ID, name = "n_genets")
```

```{r}
png(filename = "../output/cell_count_18hps_preni.png", width = 1300, height = 850)
ggplot(well_counts,
       aes(x = simple_stage,
           y = cell_count_per_well,
           color = under_tissue)) +

  geom_point(
    position = position_jitter(width = 0.08, height = 0),
    size = 9,
    alpha = 0.95
  ) +

  facet_wrap(~ treatment_mg_per_L, nrow = 1) +

scale_color_manual(
  values = c(
    "yes" = "#009E73",
    "no"  = "#CC79A7",
    "NA"  = "grey"
  ),
  labels = c(
    "yes" = "Emerging from\nunderneath tissue explant",
    "no"  = "Whole monolayer",
    "NA"  = "No monolayer"
  ),
  name = "Monolayer occurance and\narea included in cell count"
)+
  labs(
    x = "Blastogenic Stage",
    y = "Cell count per well",
  ) +

  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +

  geom_text(
    data = n_lab_genets,
    aes(x = simple_stage, y = -10, label = paste0("n = ", n_genets)),
    inherit.aes = FALSE,
    size = 7
  ) +

  coord_cartesian(clip = "off") +

  theme_classic(base_size = 18) +
  theme(
    strip.text = element_text(size = 22, face = "bold"),
    axis.title = element_text(size = 22),
    axis.text = element_text(size = 20),
    legend.title = element_text(size = 22),
    legend.text = element_text(size = 20),
    legend.position = "right",
    panel.background = element_rect(fill = "white", color = "black"),
    legend.key = element_rect(fill = "white", color = NA)
  )

dev.off()
```

```{r}
tissue_well <- cell_fact18 %>%
  distinct(well_Placements, animal_ID, treatment_mg_per_L, simple_stage, tissue_total)
tissue_summary <- tissue_well %>%
  group_by(treatment_mg_per_L, simple_stage) %>%
  summarise(
    n_wells = n(),
    mean_tissue = mean(tissue_total, na.rm = TRUE),
    sd_tissue = sd(tissue_total, na.rm = TRUE),
    .groups = "drop"
  )

tissue_summary

overall_tissue <- tissue_well %>%
  summarise(
    n_wells = n(),
    mean_tissue = mean(tissue_total, na.rm = TRUE),
    sd_tissue = sd(tissue_total, na.rm = TRUE)
  )

overall_tissue

```


